---
title: "Notes on tibble and S4Vectors::DataFrame"
author: Vince Carey
date: July 13, 2016
output: 
  html_document:
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
suppressPackageStartupMessages({
suppressWarnings({
library(BiocStyle)
library(BSgenome.Hsapiens.UCSC.hg19)
library(GenomicRanges)
})
})
})
```

# Overview

- tibble package defines utilities for working with data.frame concepts in a more convenient, streamlined way
- Bioconductor S4Vectors defines DataTable and DataFrame that are compatible with tibble concepts but are more general and accommodate genomically relevant requirements
    - run-length encoding in a column
    - DNAStringSet in a column
    - many natural coercions for use with dplyr

The key design principle is that the columns of DataFrame satisfy a simple
API, specifically

    - `length()`
    - `[`, `[<-`

# Illustration

```{r getseq}
library(BSgenome.Hsapiens.UCSC.hg19)
library(GenomicRanges)
# overlapping 8-mers
myv = Views(Hsapiens$chr1, start=100601:100800, width=8)
mytag = paste0("v", 1:200)
mygrp = Rle( rep(c(1L,2L),each=100))
dd = DataFrame(tag=mytag, grp=mygrp, octamers=myv)
dd
```

# And so on

```{r lkm}
library(dplyr)
as.data.frame(dd) %>% filter(grepl("TCAC", octamers ))
library(DT)
datatable(as.data.frame(dd))
```
